### 2021-04-22 final-report.latex/conclusion.tex
second round of Chris' notes

Using a Monte Carlo simulation, we have analyzed two main quantum field theories: the $\phi^4$ model and the $O(3)$ non-linear sigma model. To perform the calculation, we run a series of Metropolis sweeps interspersed with Wolff cluster algorithms, creating a Markov chain of samples. We use these field configurations to calculate operators using a Wick-rotated path integral.

In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.72$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality. This process verified the effectiveness of the Monte Carlo computational system. We then generalized our system to the $O(3)$ non-linear sigma model, confirming our calculations with known results. We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time. Specifically, the gradient flow reduces the linear divergence of the susceptibility to either a power-law relationship or a logarithmic relationship with $\chi^2/DOF$ of $31.3$ and $761.6$ respectively.

Finally, we produced a plot demonstrating the relationship between the topological charge and the $\theta$ in the topological action. This plot demonstrates qualitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$. In the $\tau=0$ case, this slope became vertical rapidly for higher lattice sizes while this process occurred more slowly at nonzero flow time.

Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high. This imprecision can be attributed to errors that are too small. Future work could include more independent simulations which may reduce the variance between points or increase the statistical errors. Furthermore, more and larger lattices may help reduce the $\chi^2$ value.

Procedurally, we found that the MPI parallelization and checkerboard algorithm was unnecessary in this calculation. The computation time of the Runge-Kutta algorithm for computing the gradient flow far outweighed that of the Monte Carlo simulation. A possible improvement could be a parallelization of the Runge-Kutta algorithm or a more accurate approximation to the gradient flow.

Generally, this study has implications in both condensed matter and nuclear physics as the NLSM is used in both. Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored.

Other future work may include explorations of higher order non-linear sigma models or other topological effective field theories. The latter of these models has applications in topological quantum computing and other technologies.

% MPI is unnecessary in this case

\appendix

>>>>

Using a Monte Carlo simulation, we have analyzed two main quantum field theories in 1+1 spacetime dimensions: the $\phi^4$ model and the $O(3)$ non-linear sigma model. We used the path integral formulation of QFT to simulate these quantum fields as statistical systems on a Euclidean lattice and extract information about the phase transitions and topology. To perform these calculations, we run a series of Metropolis sweeps interspersed with Wolff cluster steps, creating a Markov chain of samples. We ensure that each measurement is effectively independent by measuring the autocorrelation and thermalize the lattice to keep all measurements near the action minima.

In the $\phi^4$ model, we verified a phase transition at $m_0^2=-0.72$ using the magnitude of the average magnetization, the magnetic susceptibility, the Binder cumulant and the bimodality. This process verified the effectiveness of the Monte Carlo computational system. We then generalized our system to the $O(3)$ non-linear sigma model and transitioned to a \texttt{C++} code base. We confirmed our calculations with known results by measuring the internal energy and magnetic susceptibility. We calculated the topological susceptibility $\chi_t$ to analyze its divergence in the continuum limit, confirming that this is indeed the case even at finite flow time. Specifically, the topological susceptibility under the gradient flow follows either a power-law relationship or a logarithmic relationship as a function of lattice size, both of which diverge as $L\rightarrow\infty$.  At flow time $\tau=5t_0$, the $\chi^2/DOF$ goodness of fit values are of \powchi and \logchi respectively.

Finally, we demonstrated the relationship between the topological charge and the $\theta$ in the topological action. This plot demonstrates quantitatively the different divergences of the topological susceptibility, which is equivalent to the slope at $\theta=0$. In the $\tau=0$ case, we see the slope rapidly approach positive infinity at $\theta=0$. While this transition occurs more slowly with the application of the gradient flow, the slope continues to increase.

Though the logarithmic and power-law functions visually fit the susceptibility data well, the $\chi_2$ values are high. This imprecision can be attributed to underestimated errors. Since the $\tau=0$ case features a more acceptable fit ($\chi^2/DOF=1.8$ for the power-law), the gradient flow seemingly contributes to this error. While the Jackknife method accurately estimates statistical errors of the sample, we did not incorporate any systematic errors arising from the gradient flow calculation. Future work could reduce the tolerance of the adaptive step size algorithm to make this calculation more accurate, though this increases computational requirements substantially. Furthermore, a larger number of lattice sizes may provide a more complete picture of the continuum limit.

Procedurally, we found that the MPI parallelization and checkerboard algorithm were unnecessary in this calculation. The computation time of the Runge-Kutta algorithm for computing the gradient flow far outweighed that of the Monte Carlo simulation. A possible improvement could be a parallelization of the Runge-Kutta algorithm or a more accurate approximation to the gradient flow. Additionally, the performance of the Python simulation was slower by up to two orders of magnitude. Future studies should therefore rely solely on an efficient programming language for Monte Carlo techniques.

In the context of the long-standing question of topological suscpetibility in the NLSM, this study has diminished the plausibility that ultraviolet fluctuations cause the divergence. Instead, we are left to consider the other two options outlined in \cite{berg1981}: that the definition of the topological charge density is problematic or that the NLSM does not have a decent continuum limit. Future work could include similar numerical calculations using a different definition of the topological charge.

Generally, this study has implications in both condensed matter and nuclear physics as the NLSM is used in both. Though the calculation of the $\chi_t$ divergence confirms existing literature, the relationship between the topological charge and $\theta$ in the flow time was previously unexplored. Beyond the convergence of $\chi_t$, this relationship has applications in condensed matter where different values of $\theta$ can describe spin-chains of fermions and bosons \cite{bogli2012}. Furthermore, the mass gap has a strong relationship with the $\theta$-terms in topological NLSM, featuring massive and massless regimes \cite{allessalom2008}.

### 2021-04-22 final-report.latex/intro.tex
second round of Chris' notes

Quantum field theory (QFT) is an extraordinarily successful framework which describes a range of physical phenomena. Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics. QFT is also plays an critical role in condensed matter physics through effective field theories which model emergent effects such as phonons and quasiparticles. Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.

However this power comes at a cost: the study of quantum fields is rife with infinities. A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result. Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group which exploits the freedom of parameters such as mass and electric charge. Since these values cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results. However, there exists a set of theories which as not perturbatively renormalizable.

One such example is the \textit{non-linear sigma model} (NLSM), a prototypical theory in both condensed matter and particle physics. In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force. In general, the NLSM shares key features with non-Abelian gauge theories (such as QCD) like a mass gap and asymptotic freedom \cite{polyakov1975} and therefore is a useful model for exploring these phenomena in simpler systems.

In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time). This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space. Solutions such as these are ``topologically protected'', meaning they cannot evolve into the vacuum state via small fluctuations. Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}. Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}. In these devices, topology could protect the delicate quantum states necessary for information processing.

Since the NLSM in not perturbatively renormalizable, we require nonperturbative techniques to study topological effects. An alternative solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}. After this transformation, fields become computationally calculable. This process introduces the lattice spacing as a length scale $a$. 
% something here
Therefore, we expect any physical result to converge in the continuum limit, i.e. when $a\rightarrow 0$. However, this is not always the case as observables mix on the lattice, leading to divergences. As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.

In this study, we focus on one observable that diverges in the continuum: the topological susceptibility. The convergence of the topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{berg1981,bietenholz2018}. In QCD, mathematical techniques have proven that this quantity vanishes in the continuum limit \cite{giusti2004}, however a similar line of reasoning for the NLSM has not been successful \cite{bietenholz2018}.

@@

A second perspective on the topological susceptibility arises from the introduction of a $\theta$ term into the field Lagrangian. This term drives the vacuum state into a topological phase \cite{alles2007}. Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility. The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.

In this paper we verify the result the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the NLSM.

@@

To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov Chain Monte Carlo simulation. We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}). After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices. Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation. By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.

@@

A portion of this work began during the summer of 2021 using funding from the 1693 Scholars Program. This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model. The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honor course.

>>>>

Quantum field theory (QFT) is an framework used to describe a range of physical phenomena to a remarkable degree of accuracy. Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics. QFT also plays a critical role in condensed matter physics through effective field theories which model emergent phenomena such as phonons and quasiparticles. Compared to experiment, QFT is remarkably accurate, famously predicting the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.

However this power comes at a cost: the study of quantum fields is rife with infinities. A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result. Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group. This counter-intuitive technique exploits the freedom of parameters such as mass and electric charge. Since these constants cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results. Unfortunately, not all theories are perturbatively renormalizable.

One such example is the \textit{non-linear sigma model} (NLSM), a prototypical theory in both condensed matter and particle physics. In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force. In general, the NLSM shares key features with non-Abelian gauge theories such as QCD, including a mass gap and asymptotic freedom \cite{polyakov1975}. Therefore, the NLSM is a useful model for exploring the effect of these properties using a simpler system.

In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time). This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space. Solutions such as these are ``topologically protected,'' meaning they cannot evolve into the vacuum state via small fluctuations. Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}. Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}. In these devices, topology could protect the delicate quantum states necessary for information processing.

Since the NLSM in not perturbatively renormalizable, we require nonperturbative techniques to study topological effects. An alternative solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}. After this transformation, fields become computationally calculable. This process introduces the lattice spacing as a length scale $a$ where the \textit{continuum limit} is defined as taking $a$ to zero. We expect physical observables to converge in the continuum limit, however this is not always the case. As an example, states of definite angular momentum mix when discretized on a rectangular lattice due to a breaking of continuous rotational symmetry. This causes some operators that depend on angular momentum to suffer divergences.

In this study, we focus on one observable that diverges in the continuum: the topological susceptibility. The topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{bietenholz2018} since it is unclear if a convergent solution exists. While the some analytical arguments argue the topological susceptibility should approach zero in the continuum limit, numerical results predict infinities. In QCD, mathematical techniques proved that the susceptibility vanishes in the continuum limit \cite{giusti2004}, a fact supported by numerical calculation \cite{giusti2004}.

@@

A second perspective on the topological susceptibility arises from the introduction of a $\theta$-term into the field Lagrangian. This term drives the vacuum state into a topological phase \cite{allessalom2008}. Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility. The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.

In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.

@@

To numerically study the topological qualities of the non-linear sigma model, we first implement a Markov chain Monte Carlo simulation. We initially construct a proof-of-concept Python program that models the simpler $\phi^4$ model (see Sec.~\ref{sec:phi4}). After comparing with existing literature, we transition to a C++ simulation for efficiency, implementing the non-linear sigma model in larger lattices. Since the gradient flow has no exact solution in the non-linear sigma model, we implement a numerical solution using a fourth-order Runge-Kutta approximation with automatic step sized. By applying the gradient flow to every configuration in the sample, we can measure its effect on the topological charge and susceptibility.

@@

A portion of this work began during the summer of 2021 using funding from the 1693 Scholars Program. This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model. The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honors course.

### 2021-04-22 final-report.latex/main.tex
second round of Chris' notes

\documentclass[12pt]{report}
\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{amsfonts}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\setlength{\oddsidemargin}{0.5in}
\setlength{\evensidemargin}{0.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{.75in}
\setlength{\textheight}{8.6in}

\newcommand{\e}{\vec e}
\newcommand{\citeneeded}{\footnote{\textit{citation needed}}}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\codeexample}[1]{
    \section{\texttt{#1}}
    \lstinputlisting[language=C++]{../code/cpp_code/#1}
}

@@

I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations. And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded the summer portion of this research.

@@

\begin{abstract}
\setcounter{page}{5}
\addcontentsline{toc}{chapter}{Abstract}
\paragraph
\indent The $O(3)$ non-linear model features topological effects such as instantons but cannot be perturbatively renormalized. Other nonperturbative theories such as QCD have motivated the development of lattice quantum field theory, where fields are simulated as discrete arrays and extrapolated to the continuum limit. Unfortunately, operators such as the topological susceptibility diverge in this limit. The gradient flow is a strong contender to remove these infinites, having prior success in QCD. We confirm a previous result\cite{bietenholz2018} that the gradient flow does not make the topological susceptibility infinite. We also study the nontrivial field theory by introducing a $\theta$ term into the action and plot the topological charge as a function of $\theta$.

@@

\pagenumbering{arabic}
\chapter{Introduction}
\input{intro.tex}
\chapter{Theory}
\input{theory.tex}
\chapter{Methods}
\label{sec:methods}
\input{methods.tex}
\chapter{Results}
\input{results.tex}
\chapter{Conclusion \& Outlook}
\input{conclusion.tex}
\appendix
\chapter{C++ NLSM Monte Carlo Program}
\lstset{language=C++,
                basicstyle=\scriptsize\ttfamily,
                keywordstyle=\color{blue}\ttfamily,
                stringstyle=\color{red}\ttfamily,
                commentstyle=\color{green}\ttfamily,
                morecomment=[l][\color{magenta}]{\#},
                breaklines=true,
                postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}
}

@@

\bibliographystyle{unsrt}
\bibliography{library}

>>>>

\documentclass[12pt]{report}
\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{amsfonts}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
%\usepackage[%
    %style=numeric, 
    %sorting=none,
    %doi=false,
    %url=false,
    %giveninits=true,
    %eprint=false,
    %isbn=false,
    %clearlang=true,
    %date=year,
    %hyperref]{biblatex}
\usepackage[style=phys,biblabel=brackets,maxbibnames=3]{biblatex}
%\renewbibmacro{in:}{%
  %\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

\setlength{\oddsidemargin}{0.5in}
\setlength{\evensidemargin}{0.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{.75in}
\setlength{\textheight}{8.6in}

\newcommand{\logchi}{7.4 }
\newcommand{\powchi}{8.8 }

\addbibresource{library.bib}

\newcommand{\e}{\vec e}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\codeexample}[1]{
    \section{\texttt{#1}}
    \lstinputlisting[language=C++]{../code/cpp_code/#1}
}

@@

I also acknowledge the help the High Performance Computing group at William \& Mary, whose cluster performed many of these simulations. And finally, I would like to acknowledge the financial help of the 1693 Scholars Program which funded preliminary research over the summer.

@@

\begin{abstract}
\setcounter{page}{5}
\addcontentsline{toc}{chapter}{Abstract}
\paragraph
\indent Quantum field theory is an extraordinarily successful framework that describes phenomena in particle physics and condensed matter. The $O(3)$ non-linear sigma model (NLSM) is a specific theory used in both of these fields, describing ferromagnets and acting as a prototype for the strong nuclear force. It features topologically stable configurations known as instantons which cannot continuously evolve to the ground state. The topological susceptibility is a parameter that describes this stability and is predicted to vanish in physical theories, however numerical simulations find that the topological susceptibility diverges in the continuum limit \cite{berg1981}. This issue has motivated the application of the ``gradient flow'', a smoothing of high-frequency modes. We study the effect of the gradient flow on this divergence, finding that it persists as a logarithmic divergence. This result supports a previous study \cite{bietenholz2018} and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit. We also study the nontrivial field theory by introducing a $\theta$-term into the action and analyze the topological charge as a function of $\theta$ under the gradient flow.

@@

\pagenumbering{arabic}

\chapter{Introduction}
\input{intro.tex}
\chapter{Theory}
\input{theory.tex}
\chapter{Methods}
\label{sec:methods}
\input{methods.tex}
\chapter{Results}
\input{results.tex}
\chapter{Conclusion \& Outlook}
\input{conclusion.tex}
\appendix
\chapter{C++ NLSM Monte Carlo Program}
\lstset{language=C++,
                basicstyle=\scriptsize\ttfamily,
                keywordstyle=\color{blue}\ttfamily,
                stringstyle=\color{red}\ttfamily,
                commentstyle=\color{green}\ttfamily,
                morecomment=[l][\color{magenta}]{\#},
                breaklines=true,
                postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}
}

@@

%\bibliographystyle{unsrt}
%\bibliography{library}

\printbibliography
